A Theory of Bidirectional Type Synthesis for Simple Types

### 2023-06-26 tex/sec2-outline.tex
[ tex ] S4 draft

\begin{theorem}
A type synthesiser that uses `not in synthesising mode' as its excuse can be constructed from a mode preprocessor and a bidirectional type synthesiser.
\end{theorem}

>>>>

\begin{theorem}\label{thm:implementation}
A type synthesiser that uses `not in synthesising mode' as its excuse can be constructed from a mode preprocessor and a bidirectional type synthesiser.
\end{theorem}

### 2023-06-26 tex/sec4-pre-synthesis.tex
[ tex ] S4 draft

\begin{lemma}[Soundness]\label{thm:term-soundness}
If\/ $\Gamma \vdash_{\Sigma, \Omega} t :^\dir{d} A$, then $\Gamma \vdash_{\Sigma, \erase\Omega} t : A$.
\end{lemma}

@@

\begin{lemma}[Completeness]\label{thm:term-completeness}
If\/ $\erase\Gamma |-_{\Sigma, \Omega} t^\dir{d}$ and\/ $\Gamma |-_{\Sigma, \erase\Omega} t : A$, then $\Gamma |-_{\Sigma, \Omega} t :^\dir{d} A$.
\end{lemma}

@@

In short, soundness and completeness are no more than the separation and combination of mode and typing information carried by the three kinds of derivations while keeping their basic structure, which is directed by the same raw term.
Formulating the notion of mode derivation helps to complete this clean picture.

@@

Now we are almost ready to construct a generalised mode preprocessor.

@@

\subsection{Annotatability Is Not Completeness}
\label{sec:annotatability}

\Josh{$\frac{1}{2}$~pages}

@@

  \caption{Annotation ordering between raw terms}
  \label{fig:annotation-order}
\end{figure}

>>>>

\begin{lemma}[Soundness]\label{thm:soundness}
If\/ $\Gamma \vdash_{\Sigma, \Omega} t :^\dir{d} A$, then $\Gamma \vdash_{\Sigma, \erase\Omega} t : A$.
\end{lemma}

@@

\begin{lemma}[Completeness]\label{thm:completeness}
If\/ $\erase\Gamma |-_{\Sigma, \Omega} t^\dir{d}$ and\/ $\Gamma |-_{\Sigma, \erase\Omega} t : A$, then $\Gamma |-_{\Sigma, \Omega} t :^\dir{d} A$.
\end{lemma}

@@

In short, soundness and completeness are no more than the separation and combination of mode and typing information carried by the three kinds of derivations while keeping their basic structure, which is directed by the same raw term.
%Formulating the notion of mode derivation helps to complete this clean picture.

@@

Now we are ready to construct a generalised mode preprocessor.

@@

\subsection{Annotatability}
\label{sec:annotatability}

@@

  \caption{Annotation ordering between raw terms}
  \label{fig:annotation-ordering}
\end{figure}

\citet{Dunfield2021} formulated completeness differently from ours (\cref{thm:completeness}) and proposed \emph{annotatability} as a more suitable name.
Here we reformulate annotatability in our theory and discuss its relationship with our completeness and generalised mode preprocessing.

\begin{proposition}[Annotatability]\label{thm:annotatability}
If\/ $\Gamma |-_{\Sigma,\erase\Omega} t : A$, then there exists~$t'$ such that\/ $t' \sqsupseteq t$ and $\Gamma |-_{\Sigma,\Omega} t' :^\dir{d} A$ for some~$\dir{d}$.
\end{proposition}

Defined in \cref{fig:annotation-ordering}, the `annotation ordering' $t' \sqsupseteq t$ means that $t'$~has the same or more annotations than~$t$.
In a sense, annotatability is a reasonable form of completeness: if a term is typable in $(\Sigma, \erase\Omega)$, it may not be typable in $(\Sigma, \Omega)$ directly, but will be if some annotations are added.
Indeed, \cref{thm:annotatability} can be easily proved in our theory from generalised mode preprocessing.

\begin{proof}[Proof of \cref{thm:annotatability}]
By \cref{thm:generalised-mode-preprocessing}, there is a generalised mode derivation of $V |-_{\Sigma,\erase\Omega} t^{\dir{d}\,g\,s}$ for some mode~$\dir{d}$ and booleans $g$~and~$s$.
Perform induction on this generalised mode derivation to construct a bidirectional typing derivation in the same mode (and also the term~$t'$ and a proof of $t' \sqsupseteq t$, which are straightforward and omitted here).
\begin{itemize}
\item $\SynRule{Var}$, $\SynRule{Anno}$, and $\Rule{Op}$:
The outermost rule of the given typing derivation must be the corresponding one; apply the corresponding bidirectional typing rule to the induction hypotheses.
\item $\ChkRule{Sub}$:
Apply $\ChkRule{Sub}$ to the induction hypothesis.
\item $\SynRule{Missing}$:
This is the interesting case where we map a $\SynRule{Missing}$ rule to an $\SynRule{Anno}$ rule and add to the term a type annotation, which comes from the given typing derivation.
\vspace{-\topsep-\baselineskip}
\end{itemize}
\end{proof}

On the other hand, when using a bidirectional type synthesiser to implement type synthesis~(\cref{thm:implementation}), annotatability cannot replace our completeness since it relates typing derivations for possibly differently annotated terms, but a bidirectional type synthesiser is concerned about properties of \emph{one} given term throughout the process.
Annotatability also does not help the user to deal with missing annotations like generalised mode preprocessing does: even though annotatability seems to construct a term with missing annotations filled in, its antecedent requires a typing derivation, which is what the user is trying to construct and does not have yet.
It seems that as we focus on properties that serve practical purposes, it becomes unclear how annotatability comes into the picture.
In particular, annotatability mixes up the aspects of annotating and typing, and therefore does not fit into our theory, where the two aspects are separated for better usability.
